---
title: Using forcing functions in ordinary differential equations (ODEs) in Stan
output:
  # Use "default" placeholders below to later customize any of the output
  # formats as needed.
  html_document:
    toc: true
    number_sections: true
  pdf_document:
    default
  md_document:
    default
---

# Overview

Ordinary differential equations (ODEs)
are widely used for mathematical modeling
of natural systems.
A less common
but critical
ODE solver input
is the incorporation of fixed temporal data
representing a process whose underlying equation is unknown,
called a forcing function.
Forcing functions are more readily supported
in lower-level programming languages
than in higher-level, domain-specific programming languages (DSLs)
like Stan
that allow fewer mathematical primitives
and are more complex to extend.
The sections below
explain
how to use a forcing function
with the widely used 2-equation predator--prey ODEs example,
also called Lotka--Volterra.

# Preamble

First, load the libraries we will need for the entire exercise.

```{r load-libraries}
#| cache: false
library(rstan)
library(deSolve)
library(posterior)
library(dplyr)
library(tidyr)
library(ggplot2)
```

Data science libraries of `dplyr`, `tidyr` and `ggplot2`
require little introduction for many R users.

> **Note:**
> Those less familiar
> can learn about them
> from the freely available Carpentries lesson
> *R for Reproducible Scientific Analysis*
> (<https://swcarpentry.github.io/r-novice-gapminder/>)
> and, for a more depth, the excellent freely available book
> *R for Data Science*
> (<https://r4ds.hadley.nz/>).

While we our primarily goal is to use `rstan` to solve the ODEs,
we use `deSolve` to generate the *ground truth* data;
ground truth data here refers to
generating observations with known *true* parameters
so that we can assess how well
the fitted parameters
compare with the true parameters.

> **Note:**
> It is also possible to use Stan (`rstan`) itself
> to generate the ground truth data,
> as described in the
> Stan User's Guide >
> Example Models >
> Ordinary Differential Equations >
> Measurement error models.
> But for simplicity,
> we use R's popular `deSolve` package
> to use familiar R code where possible.

A well written Stan model
should be relatively insensitive to initial conditions.
The example below,
demostrates a case of undesirable sensitivity to initial conditions.
The `posterior` package
simplifies inspecting the data wrangling work
to visualize the ODE parameters accepted by Stan
for each chain to troubleshoot disagreement among chains.

# Typical ODEs without a forcing function

## Equations in R

The Lottka--Volterra equations are:

$$
\begin{aligned}
\frac{dx}{dt} &= \phantom{-} \alpha x - \beta x y \\
\frac{dy}{dt} &= - \gamma y + \delta x y \\
\end{aligned}
$$
where,

State variables:
$$
\begin{aligned}
x &: \text{Population of prey} \\
y &: \text{Population of predators} \\
\end{aligned}
$$
Parameters:
$$
\begin{aligned}
\alpha &: \text{Growth rate of prey} \\
\beta &: \text{Death rate of prey due to predators} \\
\gamma &: \text{Death rate of predators} \\
\delta &: \text{Growth rate of predators due to prey}
\end{aligned}
$$

To code these equations in R,
we need to wrap them in a function
with specific function arguments
common to many ODE solvers,
namely,
the time vector,
the list of state variables,
and then the list of parameters.
Therefore,
the input to `ode()` of the `deSolve` package
is written as:

```{r eqs-lv}
#| cache: true
lv <- function(t, state, params) {
  with(as.list(c(state, params)), {
    dx <-  a * x - b * x * y
    dy <- -g * y + d * x * y
    return(list(c(dx, dy)))
  })
}
```

The `with()` function above makes the contents of the lists
`state` and `params` appear in the environment
without needing to explicitly refer to them
with the `$` operator such as
`state$x` or `params$a`;
this makes the function code more legible
and more closely resemble the mathematical equations.
However,
it's generally considered bad practice in R
to use the `with()` function
outside of specific functions
that require unpacking many list arguments.

## Generate noisy input data with known ODE parameters

A common statistical practice
is to generate data with known parameters
so that we can test
how well a method or model
is able to recover the parameters used to generate the data.
Therefore,
we next simulate ODE data
using known parameters
`a`, `b`, `g`, and `d`
(i.e. $\alpha$, $\beta$, $\gamma$, and $\delta$).
Additionally,
we need to assume a known initial condition
$x_0$ and $y_0$
stored in the vector `true_y0`.
Lastly,
to make the parameter fitting
more challenging and realistic to real-world observations,
we add simulated noise.

```{r simulate-ground-truth}
#| cache: true
#| depend: -1
times <- seq(0, 15, by = 0.5)
true_y0 <- c(x = 10, y = 5.0)
true_params <- c(
  a = 1.0,
  b = .5,
  g = 1.5,
  d = 0.7
)
## Solve the ODEs, then add noise to each variable using rnorm.
pristine <- ode(
  y = true_y0,
  times = times,
  func = lv,
  parms = true_params)
noisy <- pristine
set.seed(123)
noisy[, -1] <-
  pristine[, -1] + rnorm(nrow(pristine) * (ncol(pristine) - 1),
                         sd = 0.5)
noisy[noisy < 0] <- 0
```

Using `ode()` above only solves the system of equations
at the timepoints specified by `times`.
The simulated "pristine" and "noisy" data points are shown below,
although from here on we will only use the noisy simulated data.

```{r plot-noise-added-to-ground-truth}
#| cache: true
#| dependson: -1
bind_rows(pristine = as.data.frame(pristine),
          noisy = as.data.frame(noisy),
          .id = "type") |>
  pivot_longer(-c(type:time), names_to = "var") |>
  mutate(type = factor(type, levels = c("pristine", "noisy"))) |>
  ggplot(aes(time, value, color = var)) +
  theme_bw() +
  facet_wrap(~ type, ncol = 1) +
  geom_line() +
  geom_point()
```

## Fit the ODE parameters using the equivalent Stan model

The model we are going to fit
is saved in a separate `lv.stan` file
that looks like this:
<!--
The chunk below prints the lv.stan file.
We don't actually need to store the model into `output.var`
and the value of the output.var is never used,
but we will get an error if we don't store into `output.var`.
-->

```{stan lv.stan, code = readLines("lv.stan"), output.var = "lv.stan", cache = TRUE}
```

Call the `stan()` function
to supply the inputs
to the model described above in the `lv.stan` file:

```{r simulate-stan}
#| cache: true
#| dependson: -3
fit1 <- stan(
  model_name = "lotka-volterra",
  chains = 4,
  cores = 4,
  warmup = 1000,
  iter = 5000,
  file = "lv.stan",
  data = list(
    T = nrow(noisy) - 1,
    y = noisy[-1, c("x", "y")],
    y0 = true_y0,
    t0 = 0,
    ts = noisy[-1, "time"],
    g = true_params["g"],
    d = true_params["d"]),
  seed = 12345
)
```

We get warnings
about how our initial conditions
were not able to agree
and mix together
that we will explore in the next section.

## Inspect the Stan fit

```{r tab-results}
#| cache: true
#| dependson: -1
## Numerical summary of the fits.
fit1

## Parameter plot of the fits.
plot(fit1)

## Find initial values creating problems with chains converging.
traceplot(fit1, inc_warmup = TRUE, window = c(0, 200))

## Show all samples including warmup.
traceplot(fit1, inc_warmup = TRUE)

## Show the pairs plot.
pairs(fit1, pars = c("a", "b", "lp__"))

## Plot ODEs using a sample of 100 parameters from each of the chains.
n <- 100
set.seed(123)
posterior::as_draws_df(fit1) |>
  select(.chain, .iteration, a, b) |>
  group_by(.chain) |>
  ## Sample greater variation by only choosing one sample from each ntile.
  mutate(qa = ntile(a, n / 5),
         qb = ntile(b, n / 5)) |>
  group_by(.chain, qa, qb) |>
  sample_n(1) |>
  ## Now sample the 100 parameters from each chain.
  group_by(.chain) |>
  sample_n(n) |>
  ungroup() |>
  rowwise() |>
  mutate(mat = list(
    ode(
      y = true_y0,
      times = times,
      func = lv,
      parms = list(
        a = a,
        b = b,
        g = true_params["g"],
        d = true_params["d"])) |>
      as.data.frame())) |>
  unnest(cols = "mat") |>
  ungroup() |>
  pivot_longer(-c(.chain:time), names_to = "var") |>
  mutate(.chain = as.factor(.chain)) |>
  ggplot(aes(x = time,
             y = value,
             group = interaction(.chain, .iteration, var),
             color = var)) +
  theme_bw() +
  facet_wrap(~ .chain, labeller = "label_both") +
  geom_line()
```

In the last plot above
that shows 100 ODE samples simulated by each of the 4 chains,
we see that one of the chains
is not able to capture the true periodicity.
This is because of two reasons:
(1) the oscillatory ODE system contains "peaks"
of higher posterior probabilities
and moving between these "peaks" is hard
and (2) we are using large priors
`a ~ normal(0.5, 0.5)` and `b ~ normal(0.5, 0.5)`:

```{r}
#| cache: true
## Plot the priors for `a` and `b` used in the Stan model.
data.frame(x = seq(0, 10, .1)) |>
  ggplot(aes(x)) +
  theme_bw() +
  stat_function(fun = dlnorm,
                args = c(meanlog = 0.5,
                         sdlog = 0.5))
```

But the
bottom row of the
pairs plot
further above
shows that the values of `b` ($\beta$)
around 0.5 have the highest `lp__` (log-probability)
and that values of `a` ($\alpha$)
around 1.0 have the highest `lp__`;
0.5 and 1.0 are indeed their true values.
We often learn more when things go "wrong"
and hopefully reading these diagnostic plots
has helped you learn a little more about troubleshooting
models that do not perfectly fit the noisy data!

# ODE with a forcing function

## Modified equations in R

Finally!
We get to discuss the main topic about
how to use a forcing function
when the equation of a state variable is unknown.
Let's say that we did not know
the equation terms for the predator state variable, `y`.
Then, instead of two ODEs, we need to solve a single ODE.
But we must replace the `y` term in the `x` equation
with a fixed trajectory of `y`.

Before we had:

$$
\begin{aligned}
\frac{dx}{dt} &= \phantom{-} \alpha x - \beta x y \\
\frac{dy}{dt} &= - \gamma y + \delta x y \\
\end{aligned}
$$

Now we have only the first equation and
pretend we do not yet know the process that governs the second equation:

$$
\frac{dx}{dt} = \phantom{-} \alpha x - \beta x y \\
$$

We cannot fix `y` as a single value;
`y` must vary in some way because `x` varies
and `y` and `x` influence eachother.
So we must supply a fixed trajectory of `y`
based on the noisy data
that only depends on time
because `y` is no longer part of the system of equations
that the ODE solver has access to.
We must define an external function
that produces a fixed trajectory of `y`.

Ideally,
we would use a spline function.
Splines are wonderful mathematical funtions
because they're differentiable
while still allowing us to fit complex, squiggly looking curves.
However,
spline functions are
numerically harder to implement
and Stan does not yet provide the mathematical primitives
to easily add them.
Therefore,
we will use a simpler polynomial function:

$$
\begin{aligned}
\frac{dy}{dt} &= c_n t^n + c_{n-1} t^{n-1} \ldots + c_2 t^2 + c_1 t^1 + c_0 t^0 \\
\implies
\frac{dy}{dt} &= \sum_{i=0}^n c_i t^i \\
\end{aligned}
$$

Where,
the polynomial terms are $t_0, t_1, t_2, \ldots, t_{n-1}, t_n$
and their corresponding coefficients are $c_0, c_1, c_2 \ldots, c_{n-1}, c_n$.


> **Note:**
> Unlike a spline function,
> a polynomial function has trouble fitting periodic data.
> See the plot below.

```{r poly-choose}
#| cache: true
#| dependson: "simulate-ground-truth"
## Plot different degrees of polynomials fit to the prey data.
max_time <- 7
max_degree <- length(times[times <= max_time]) - 2
fit_poly <- list()
for (i in seq_len(max_degree)) {
  fit_poly[[i]] <-
    lm(y ~ poly(time, i),
       data =
         noisy[, c("time", "y")] |>
         as.data.frame() |>
         filter(time <= max_time))
}
noisy |>
  as.data.frame() |>
  filter(time <= max_time) |>
  mutate(across(time,
                setNames(lapply(seq_len(max_degree),
                                function(i) {
                                  return(function(time) predict(fit_poly[[i]]))
                                }),
                         paste0("y_poly_", seq_len(max_degree))))) |>
  pivot_longer(-c(time:y), names_to = "degree") |>
  mutate(degree =
           sub("time_y_poly_", "", degree) |>
           as.integer()) |>
  ggplot(aes(time, value)) +
  facet_wrap(~ degree) +
  geom_line() +
  geom_point(aes(y = y))

## T-test showing other than the intercept, terms up to and including degree
## that captures the most variance.
summary(fit_poly[[length(fit_poly)]])

## Choose our polynomial degree based on the first p.value elbow.
degrees <-
  summary(fit_poly[[length(fit_poly)]]) |>
  broom::tidy() |>
  filter(term != "(Intercept)") |>
  mutate(p_value_cumsum = cumsum(p.value),
         degree = gsub(term, pattern = ".*[)]", replacement = "") |>
           as.integer())
degree <- 3
degrees |>
  ggplot(aes(x = degree, y = p_value_cumsum)) +
  geom_line() +
  geom_vline(xintercept = degree, linetype = 2)
## Workaround bug in non-standard evaluation of predict.lm().
i <- degree
```

```{r eqs-lv-ff}
#| cache: true
#| dependson: -1
## Now generate the ODE data using the forcing function (using t) and
## ignore any input for predator (dy).
y <- function(t) {
  predator <- predict(fit_poly[[degree]], data.frame(time = t))
  if (predator < 0) {
    return(0)
  }
  predator
}
lv_poly <- function(t, state, params) {
  with(as.list(c(state, params)), {
    ## Save computation time by not recalculating y(t) everytime.
    y_ <- y(t)
    dx <- a * x - b * x * y_
    return(list(c(dx)))
  })
}
```

```{r plot-lv-ff}
#| cache: true
#| dependson: -1
fit_ode_ff <- ode(
  y = noisy[1, "x"],
  times = times[times <= max_time],
  func = lv_poly,
  parms = true_params[c("a", "b")])
noisy |>
  as.data.frame() |>
  filter(times <= max_time) |>
  rename_with(~ paste0(.x, "_orig"), -time) |>
  bind_cols(x_ff = fit_ode_ff[, 2]) |>
  bind_cols(y_ff = sapply(times[times <= max_time], y)) |>
  pivot_longer(-time) |>
  separate_wider_delim(name, delim = "_", names = c("name", "type")) |>
  mutate(type = factor(type, levels = c("orig", "ff"))) |>
  ggplot(aes(time, value, group = name, color = name)) +
  facet_wrap(~ type) +
  geom_line() +
  geom_point(data = . %>%
               filter(type == "orig") %>%
               mutate(type = factor("ff", levels = c("orig", "ff"))))
```

## Fit the ODE parameters using the equivalent modified Stan model

We implement these equations in the Stan code.

The model we are going to fit
is saved in a separate `lv-forcing-function.stan` file
that looks like this:
<!--
The chunk below prints the lv.stan file.
We don't actually need to store the model into `output.var`
and the value of the output.var is never used,
but we will get an error if we don't store into `output.var`.
-->

```{stan lv-ff.stan, code = readLines("lv-forcing-function.stan"), output.var = "lvff.stan", cache = TRUE}
```

```{r simulate-stan-forcing-function}
#| cache: true
#| dependson: -3
## Run Stan using the same forcing function.
fit_stan <-
  stan(
    model_name = "lotka-volterrs-forcing-function",
    chains = 4,
    cores = 4,
    warmup = 1000,
    iter = 5000,
    file = "lv-forcing-function.stan",
    data = list(
      ## Must remove the row containing timepoint 0 and instead
      ## provide it with the initial conditions.
      ##
      ## Dimensions:
      T = sum(times > 0 & times <= max_time),
      V = ncol(noisy) - 2L,
      ## Experimental data:
      ts = times[times > 0 & times <= max_time],
      y = noisy[which(times > 0 & times <= max_time), "x", drop = FALSE],
      ## Coefficients of fitted polynomial:
      N_coef = length(coef(fit_poly[[degree]])),
      coef = coef(fit_poly[[degree]]),
      ## Initial conditions:
      t0 = 0,
      y0 = as.array(noisy[1, "x"])),
    seed = 123
)
```

## Inspect the Stan forcing function fit

```{r tab-results-forcing-function}
#| cache: true
#| dependson: -1
fit_stan

plot(fit_stan)

traceplot(fit_stan)

traceplot(fit_stan, inc_warmup = TRUE, window = c(0, 200))

pairs(fit_stan, pars = c("a", "b", "lp__"))

## Plot ODEs using a sample of 100 parameters from each of the chains.
set.seed(123) # For sample_n below.
## Running the ODEs here takes about minute to run (4 chains x 100 samples x
## 70 timepoints).
n <- 100
system.time({
  df_stan_fit <-
    posterior::as_draws_df(fit_stan) |>
    select(.chain, .iteration, a, b) |>
    group_by(.chain) |>
    ## Sample greater variation by only choosing one sample from each ntile.
    mutate(qa = ntile(a, n / 4),
           qb = ntile(b, n / 4)) |>
    group_by(.chain, qa, qb) |>
    sample_n(1) |>
    ## Now sample the 100 parameters from each chain.
    group_by(.chain) |>
    sample_n(n) |>
    ungroup() |>
    rowwise() |>
    mutate(mat = list(
      ode(
        y = true_y0[1],
        times = times[times <= max_time],
        func = lv_poly,
        parms = list(
          a = a,
          b = b)) |>
        as.data.frame())) |>
    unnest(cols = "mat") |>
    rowwise() |>
    mutate(y = y(time)) |>
    ungroup() |>
    pivot_longer(-c(.chain:time), names_to = "var") |>
    mutate(.chain = as.factor(.chain))
})

## Plot the fitted ODEs from Stan.
df_stan_fit |>
  ggplot(aes(x = time,
             y = value,
             group = interaction(.chain, .iteration, var),
             color = var)) +
  theme_bw() +
  facet_wrap(~ .chain) +
  geom_line() +
  ## Show the predator data provided to Stan.
  geom_point(
    data =
      noisy[, c("time", "x")] |>
      as.data.frame() |>
      pivot_longer(-time),
    inherit.aes = FALSE,
    aes(x = time, y = value),

    size = 0.1)
```

# Finishing thoughts

1. Don't simply choose a forcing function only based
   on how it fits its own data;
   instead,
   choose a forcing function
   by how well the other equations fit their data
   by running the full ODE system.

1. Spline functions are great as forcing functions,
   but not always practical to use.
   Polynomials can work,
   but may not work well with some types of periodic data;
   subset the data to timepoints
   where the fit looks good.

1. You may have noticed `set.seed(...)` used in a few places.
   This is to make the stochastic code outputs reproducible,
   but one would typically not use these;
   it is more common to use store the value of `.Random.seed`
   to record a seed used for a simulation
   than to force a specific outcome
   because well-adjusted simulations should produce
   very similar results
   in spite of pseudo-random number generator initial states.

# References
